---
title: "Magnitud response Mixed Model"
author: "Abraham Arbelaez and Sofia Cominelli"
date: "2025-11-21"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(readxl)
library(dplyr)
library(janitor)
library(ggrepel)
library(lme4)
library(DHARMa)
library(emmeans)
library(purrr)
```

## Read file
```{r}
df <- read_excel("P_kansas_wheat.xlsx") %>%
  clean_names()

```

# Data wrangling
```{r}
df_wheat <- df %>%
  select(year, loc_n, rep, plot, trt, source, p_kg, p_m_ppm, hg_kg_ha) %>%
  filter(source %in% c("MAP", "Check")) %>%
  rename(location = loc_n,
         rate = p_kg,
         soil_P = p_m_ppm,
         yield = hg_kg_ha) %>%
  filter(!is.na(yield))


df_wheat <- df_wheat %>%
  mutate(location = factor(location),
    rep= factor(rep),
    rate= factor(rate),
    rate_num = as.numeric(as.character(rate)))

```

# Model
The dataset analyzed in this study has 24 site years in which wheat grain yield was evaluated across increasing phosphorus fertilizer rates (P rate). The experimental structure follows a one-way factorial design, where the factor P rate includes four levels (0, 45, 90, and 135 kg P per hectare). At each site year, the experiment was arranged as a randomized complete block design with four replications. Within each replication, all four P rate treatments were applied, providing a balanced RCBD structure for analysis.

The experiment model can be writen as: 
$$ y_{ijk} = \mu + p_i+ s_j + b_{k(j)}  + \epsilon_{ijk}$$
where:
$y_{ijk}$ is the wheat grain yield of the $i$th level
$\mu$ is the overall mean
$pi_i$ is the treatment mean for the $i$th treatment ($i$= 1,2,3,4)
$s_k$ is the random effect of the $k$th site-year ($k$= 1, 2,...24)
$b_{j(k)}$ is the random effect of the $j$th block in the $k$th site-year ($j$= 1,2,3,4)
$\epsilon_{ijk}$ is the residual error for the corresponding $i$th treatment, $j$th block, and $k$th site-year

 $$S_j \sim N(0, \sigma^2_s)$$  $$b_{k(j)}\sim N(0, \sigma^2_b)$$ $$ \varepsilon_{ijk} \sim N(0,\sigma^2_e),$$.

```{r}
model_1 <- lmer(yield ~ rate + (1 | location), data = df_wheat)
model_1

```

```{r}
plot(simulateResiduals(model_1))
```

# Marginal means
```{r}
emm_1 <- emmeans(model_1, ~ rate)
emm_1

plot(emm_1)
```
## Site level means by rate
```{r}
site_rate_means <- df_wheat %>%
  group_by(year, location, rate) %>%
  summarise(mean_yield = mean(yield, na.rm = TRUE), .groups = "drop")

site_means_wide <- site_rate_means %>%
  mutate(rate_num = as.numeric(as.character(rate))) %>%
  select(year, location, rate_num, mean_yield) %>%
  tidyr::pivot_wider(
    names_from  = rate_num,
    values_from = mean_yield,
    names_glue  = "mean_yield_{rate_num}"
  ) %>%
  mutate(control_yield= mean_yield_0,
    max_yield= pmax(
      mean_yield_0,
      mean_yield_45,
      mean_yield_90,
      mean_yield_135,
      na.rm = TRUE),
    delta_yield= max_yield - control_yield,
    relative_yield= dplyr::if_else(is.finite(max_yield) & max_yield > 0,100 * control_yield / max_yield, NA_real_))


```

## Linear slope, intercept, STP per site year
```{r}
df_slope_site <- df_wheat %>%
  group_by(year, location) %>%
  nest() %>%
  mutate(int_slope = map(data, ~ lm(yield ~ rate_num, data = .x)),
    intercept = map_dbl(int_slope, ~ coef(.x)[["(Intercept)"]]),
    slope= map_dbl(int_slope, ~ coef(.x)[["rate_num"]])) %>%
  ungroup() %>%
  select(year, location, intercept, slope)

stp_site <- df_wheat %>%
  group_by(year, location) %>%
  summarise(STP = mean(soil_P, na.rm = TRUE), .groups = "drop")

df_slope_site <- df_slope_site %>%
  left_join(stp_site, by = c("year", "location")) %>%
  mutate(site_id = paste0(year, "_", location))

p_max <- max(df_wheat$rate_num, na.rm = TRUE)

df_features_site <- df_slope_site %>%
  mutate(delta_y = slope * p_max) %>%
  select(year, location, site_id, intercept, slope, STP, delta_y)
```

## Site info table
```{r}
site_metrics <- site_means_wide %>%
  left_join(df_features_site, by = c("year", "location"))

```


